---
title: "covid_project"
author: "Xiaohan Xie"
format: html
editor: visual
self-contained: true
execute:
  warning: false
---

```{r include=FALSE}
library(HiCExperiment)
library(HiContacts)    
library(GenomicRanges)
library(InteractionSet)
library(rtracklayer)
library(tidyverse)
library(epiwraps)
library(patchwork)
library(GenomicFeatures)
```

## 1. Load Hi-C Files
Using the command line tool of cooler, we merged the replicates for Mock and SARS-CoV-2-infected conditions. In accordance with the original study's methods, we processed the data at 10kb resolution. We then applied matrix balancing normalisation to correct for biase in the Hi-C reads. Lastly, we performed matrix coarsening to generate multi-resolution data. 

We limited our analysis to chromosome 14 to reduce the computational complexity. 
```{bash, eval=FALSE}
cooler merge merged_mock_HiC.mcool mock_HiC_rep1.mcool::/resolutions/10000 mock_HiC_rep2.mcool::/resolutions/10000
cooler merge merged_cov_HiC.mcool cov_HiC_rep1.mcool::/resolutions/10000 cov_HiC_rep2.mcool::/resolutions/10000
cooler balance merged_mock_HiC.mcool
cooler balance merged_cov_HiC.mcool
cooler zoomify -r 10000,200000 merged_mock_HiC.mcool -o merged_mock_HiC_multi.mcool
cooler zoomify -r 10000,200000 merged_cov_HiC.mcool -o merged_cov_HiC_multi.mcool
```

```{r}
hic_mock_merged <- HiCExperiment("./GSE179184_RAW/merged_mock_HiC_multi.mcool", focus = "chr14", resolution = 10000)
hic_cov_merged <- HiCExperiment("./GSE179184_RAW/merged_cov_HiC_multi.mcool", focus = "chr14", resolution = 10000)
length(hic_mock_merged)
length(hic_cov_merged)
summary(interactions(hic_mock_merged)$balanced)
summary(interactions(hic_cov_merged)$balanced)
```
Mock sample has slightly more interaction counts than the CoV-infected group. Both samples show highly skewed distribution, as the majority of interactions tend to have very balanced scores.

## 2. Insulation Scores and Boundary strengths
According to methods decribed in the original paper, we calculated the insulation scores using a 200kb sliding window. TAD boundaries were then called by finding the local minima in the insulation profile with a threshold of 0.5. 

```{r, warning=FALSE}
hic_mock <- getDiamondInsulation(hic_mock_merged, window_size = 200000)
hic_cov <- getDiamondInsulation(hic_cov_merged, window_size = 200000)
hic_mock <- getBorders(hic_mock, strong_threshold = 0.5)
hic_cov <- getBorders(hic_cov, strong_threshold = 0.5)
bounds_mock <- topologicalFeatures(hic_mock, 'borders')
bounds_cov <- topologicalFeatures(hic_cov, 'borders')
```
In order to compare between two conditions, we merged the detected boundaries from both Mock and CoV groups. 
```{r}
merged_bounds <- GenomicRanges::reduce(c(bounds_mock, bounds_cov))
merged_bounds <- merged_bounds[seqnames(merged_bounds) == "chr14"]
head(merged_bounds)
ins_track_mock <- metadata(hic_mock)$insulation
ins_track_cov <- metadata(hic_cov)$insulation
```
We then used the epiwraps package to plot the distribution of insulation scores, extending 500kb upstream and downstream. 

```{r, fig.cap="Insulation scores calculated TAD boundaries in both conditions."}
ins_track_mock$score <- ins_track_mock$insulation
ins_track_cov$score  <- ins_track_cov$insulation
tracks <- list(Mock = ins_track_mock, SARS_CoV_2 = ins_track_cov)
ins <- signal2Matrix(tracks, regions = merged_bounds, extend = 500000, w = 10000)
plotEnrichedHeatmaps(ins, scale_title ="Insulation Scores", use_raster = FALSE)
```

We can see that although CoV-infected group shows a slightly higher insulation score at the consensus boundaries, this effect is rather mild, as reported in the original study. 

## 3. TADs and Intra-TAD Interaction
We then further obtained the TADs, defined as genomic intervals nested between the previously identified consensus boundaries. 
```{r}
# gaps between merged boundaries are TADs
merged_tads <- gaps(merged_bounds)
merged_tads <- merged_tads[strand(merged_tads) == "*"] 
merged_tads <- merged_tads[seqnames(merged_tads) == "chr14"]
# filter out extremely small fragments (as we have resolution of 10000, use threshold 20000)
merged_tads <- merged_tads[width(merged_tads) >= 20000]
# get the GInteractions objects
gi_mock <- interactions(hic_mock)
gi_cov <- interactions(hic_cov)
# filter out very short distant interactions (< 15 kb)
gi_mock_sub <- gi_mock[pairdist(gi_mock) >= 15000]
gi_cov_sub <- gi_cov[pairdist(gi_cov) >= 15000]
```

To quantify the intra-TAD interactions, we defined a function to isolate interactions that occur strictly within the same TAD domain. We then calculated the mean intra-TAD contacts for both groups. 
```{r}
# func to map interactions inside TADs
get_intra_tad_means <- function(gi_filtered, tads) {
  
  hits1 <- findOverlaps(anchors(gi_filtered, type="first"), tads)
  hits2 <- findOverlaps(anchors(gi_filtered, type="second"), tads)
  
  # keep only interactions where both anchors found in the same TAD
  common_hits <- inner_join(
    as.data.frame(hits1), 
    as.data.frame(hits2), 
    by = "queryHits", suffix = c("1", "2")
  ) %>% filter(subjectHits1 == subjectHits2)
  
  # aggregate mean contact score per TAD

  df <- data.frame(
    tad_idx = common_hits$subjectHits1,
    contact_val = gi_filtered$balanced[common_hits$queryHits] 
  )
  
  df_mean <- df %>% 
    group_by(tad_idx) %>% 
    summarize(mean_contact = mean(contact_val, na.rm = TRUE))
  
  return(df_mean)
}

mock_means <- get_intra_tad_means(gi_mock_sub, merged_tads)
cov_means <- get_intra_tad_means(gi_cov_sub, merged_tads)
```

```{r}

tad_metrics <- data.frame(tad_id = 1:length(merged_tads)) %>%
  left_join(mock_means, by = c("tad_id" = "tad_idx")) %>%
  rename(Mock_Mean = mean_contact) %>%
  left_join(cov_means, by = c("tad_id" = "tad_idx")) %>%
  rename(Cov_Mean = mean_contact) %>%
  mutate(
    log2FC = log2((Cov_Mean) / (Mock_Mean))   
  )

# save it as a col
mcols(merged_tads) <- tad_metrics 

```

```{r, fig.width=3, fig.height=4, fig.cap="log2FCs of intra-TAD Hi-C interactions."}

# Fig.4c: boxplot of intra-TAD interaction changes
ggplot(tad_metrics, aes(x = "Infection", y = log2FC)) +
  geom_violin(fill = "#E74C3C", width = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(
    x = "",
    y = expression(log[2]~"Fold Change (SARS-CoV-2 / Mock)")
  ) +
  theme_classic(base_size = 14)
```
It can be seen from the plot that SARS-CoV-2 infection significantly reduces the intra-TAD contacts, which aligns with the findings of the paper. 
## 4. Hi-C Matrices at a Specific TAD Domain 
As a full Aggregate Domain Analysis (ADA) proved to be computationally demanding, following the original study, we defined a specific TAD domain of interest on chromosome 14 and visualised Hi-C contacts at this region. We proceeded with bin size of 10kb, unlike in the paper, where the authors switched to a resolution of 5kb for this analysis. We calculated a differential contact matrix for this TAD region, which we visualised together with the contact matrices of both Mock and CoV groups.
```{r}

# the target genomic coordinates
target_chr <- "chr14"
target_start <- 71400000
target_end   <- 72600000
target_region <- paste0(target_chr, ":", target_start, "-", target_end)
target_gr <- GRanges(seqnames = target_chr, ranges = IRanges(target_start, target_end))

hic_mock_reg <- refocus(hic_mock, target_region)
hic_cov_reg <- refocus(hic_cov,  target_region)

hic_div      <- divide(hic_cov_reg, by = hic_mock_reg, use.scores = "balanced")

```

```{r, fig.width=9, fig.height=4, fig.cap="Hi-C matrices (bin size = 10 kb) at the indicated region (hg19) and log2FC after infection."}

# mock
p_matrix_mock <- plotMatrix(hic_mock_reg, use.scores = "balanced", scale = "log10", caption = FALSE) +
  labs(title = "Mock", fill = "log10 Score", y = paste0(target_chr, ": 71,400,000-72,600,000")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

# cov
p_matrix_cov <- plotMatrix(hic_cov_reg, use.scores = "balanced", scale = "log10", caption = FALSE) +
  labs(title = "SARS-CoV-2", fill = "log10 Score") +
  theme(axis.title.y = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())

# log2 FC
p_matrix_l2fc <- plotMatrix(
  hic_div, 
  use.scores = "balanced.l2fc",  
  scale = "linear",             
  cmap = bgrColors(), caption = FALSE, 
  limits = c(-2, 2)
) +
  labs(title = "SARS-CoV-2/mock", fill = "log2 FC") +
  theme(axis.title.y = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())

matrix_panels <- p_matrix_mock + p_matrix_cov + p_matrix_l2fc + 
  plot_layout(ncol = 3)

matrix_panels
```
The patterns here indicate that while intra-TAD interactions tend to be reduced by infection of SARS-CoV-2, contacts outside of TADs are largely unaltered or even increased. Looking at the contact matrices of the two groups (Mock and CoV), we can also see that TAD-boundaries are noticeably perturbed by infection, complementing the results of the insulation score analysis. 
